---
title: "Gonorrhea Surveillance and Trend Analysis, United States 2000–2022"
subtitle: "02 — Stochastic SIS Model Simulation"
author: "[Your Name]"
date: "2026-03-03"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    highlight: tango
    df_print: kable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width  = 9,
  fig.height = 5.5,
  dpi     = 300
)
```

## Background

Gonorrhea is modeled as an **SIS** (Susceptible–Infectious–Susceptible) process,
reflecting the absence of lasting immunity after recovery. The basic reproduction
number is defined as $R_0 = \beta / \gamma$. Parameters are calibrated
approximately to observed US gonorrhea dynamics.

## 1. Load Libraries

```{r libraries}
library(ggplot2)
library(dplyr)
library(tidyr)
library(purrr)
library(readr)

set.seed(6270)  # For reproducibility
```

## 2. SIS Model Function

```{r sis-function}
#' Stochastic SIS model for gonorrhea
#'
#' @param N       Total population size
#' @param I0      Initial number infectious
#' @param beta    Transmission rate (contacts/week × P(transmission))
#' @param gamma   Recovery rate (1 / mean duration in weeks)
#' @param t_max   Number of time steps (weeks)
#' @param dt      Time step size (default 1 week)
#' @return A data frame: time, S, I, S_prop, I_prop
run_sis <- function(N = 100000, I0 = 500, beta, gamma,
                    t_max = 260, dt = 1) {

  S <- numeric(t_max + 1)
  I <- numeric(t_max + 1)

  S[1] <- N - I0
  I[1] <- I0

  for (t in 1:t_max) {
    lambda         <- beta * I[t] / N
    new_infected   <- rbinom(1, S[t], prob = 1 - exp(-lambda * dt))
    new_recovered  <- rbinom(1, I[t], prob = 1 - exp(-gamma * dt))
    S[t + 1] <- max(0, S[t] - new_infected + new_recovered)
    I[t + 1] <- max(0, I[t] + new_infected - new_recovered)
  }

  data.frame(
    time   = 0:t_max,
    S      = S,
    I      = I,
    S_prop = S / N,
    I_prop = I / N
  )
}
```

## 3. Define Scenarios

Gonorrhea natural history assumptions:

- **Mean duration of infection:** ~5 weeks → $\gamma = 0.20$/week  
- **$R_0$** estimated at 1.5–2.5 in endemic settings

| Scenario | $\beta$ | $\gamma$ | $R_0$ | Description |
|---|---|---|---|---|
| Baseline | 0.36 | 0.20 | 1.8 | Endemic equilibrium |
| High Transmission | 0.56 | 0.20 | 2.8 | Higher-risk network / untreated population |
| Intervention | 0.22 | 0.20 | 1.1 | 50% reduction via contact tracing + treatment |

```{r define-scenarios}
scenarios <- list(
  baseline     = list(beta = 0.36, gamma = 0.20, label = "Baseline (R0=1.8)"),
  high_beta    = list(beta = 0.56, gamma = 0.20, label = "High Transmission (R0=2.8)"),
  intervention = list(beta = 0.22, gamma = 0.20, label = "Intervention (R0=1.1)")
)

n_sims <- 100     # Stochastic runs per scenario
N      <- 100000
I0     <- 500
t_max  <- 260     # 5 years in weeks
```

## 4. Run Simulations

```{r run-simulations}
all_results <- map_dfr(names(scenarios), function(sc_name) {
  sc <- scenarios[[sc_name]]
  R0 <- sc$beta / sc$gamma

  cat(sprintf("  Scenario: %-12s | beta=%.2f | gamma=%.2f | R0=%.2f\n",
              sc_name, sc$beta, sc$gamma, R0))

  map_dfr(1:n_sims, function(sim_id) {
    run_sis(N = N, I0 = I0,
            beta  = sc$beta,
            gamma = sc$gamma,
            t_max = t_max) %>%
      mutate(
        sim_id   = sim_id,
        scenario = sc_name,
        label    = sc$label,
        beta     = sc$beta,
        gamma    = sc$gamma,
        R0       = R0
      )
  })
})

cat(sprintf("Total rows generated: %d\n", nrow(all_results)))
```

## 5. Summarize Results

```{r summarize}
sim_summary <- all_results %>%
  group_by(scenario, label, time, beta, gamma, R0) %>%
  summarise(
    mean_I  = mean(I_prop),
    lower_I = quantile(I_prop, 0.025),
    upper_I = quantile(I_prop, 0.975),
    .groups = "drop"
  )
```

### Equilibrium Prevalence at Week 260

```{r equilibrium-table}
sim_summary %>%
  filter(time == t_max) %>%
  mutate(
    mean_pct  = round(mean_I  * 100, 2),
    lower_pct = round(lower_I * 100, 2),
    upper_pct = round(upper_I * 100, 2)
  ) %>%
  select(scenario, R0, mean_pct, lower_pct, upper_pct)
```

```{r save-sim-data, include=FALSE}
write_csv(
  all_results %>% select(sim_id, time, S, I, S_prop, I_prop, scenario, beta, gamma, R0),
  "gonorrhea_simulated.csv"
)
```

## 6. Visualize Results

```{r color-labels, include=FALSE}
scenario_colors <- c(
  "baseline"     = "#2166ac",
  "high_beta"    = "#d6604d",
  "intervention" = "#4dac26"
)

scenario_labels <- c(
  "baseline"     = "Baseline (R0 = 1.8)",
  "high_beta"    = "High Transmission (R0 = 2.8)",
  "intervention" = "Intervention (R0 = 1.1)"
)
```

### Figure 1: Simulation Ribbon Plot

```{r fig1-sim-ribbon}
p_sim <- ggplot(sim_summary,
                aes(x     = time / 52,
                    y     = mean_I * 100,
                    color = scenario,
                    fill  = scenario)) +
  geom_ribbon(aes(ymin = lower_I * 100, ymax = upper_I * 100),
              alpha = 0.2, color = NA) +
  geom_line(linewidth = 1.1) +
  scale_color_manual(values = scenario_colors, labels = scenario_labels, name = "Scenario") +
  scale_fill_manual( values = scenario_colors, labels = scenario_labels, name = "Scenario") +
  scale_x_continuous(breaks = 0:5) +
  labs(
    title    = "Stochastic SIS Model: Gonorrhea Transmission Scenarios",
    subtitle = "Mean % infectious ± 95% simulation interval (n=100 runs per scenario)",
    x        = "Time (Years)",
    y        = "% Population Infectious",
    caption  = paste0("N = ", format(N, big.mark = ","),
                      "; Initial infectious = ", I0,
                      "; gamma = 0.20/week (mean duration ~5 weeks)")
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title       = element_text(face = "bold"),
    legend.position  = "bottom",
    panel.grid.minor = element_blank()
  )

p_sim
ggsave("output/figures/simulation_outcomes.png", plot = p_sim,
       width = 9, height = 5.5, dpi = 300)
```

### Figure 2: R0 Sensitivity — Equilibrium Prevalence

```{r fig2-r0-sensitivity, fig.width=8, fig.height=5}
r0_vals <- seq(1.0, 3.0, by = 0.1)
eq_prev <- ifelse(r0_vals > 1, (1 - 1 / r0_vals) * 100, 0)
r0_df   <- data.frame(R0 = r0_vals, eq_prevalence = eq_prev)

sc_points <- data.frame(
  R0            = c(1.8, 2.8, 1.1),
  eq_prevalence = c((1 - 1/1.8)*100, (1 - 1/2.8)*100, (1 - 1/1.1)*100),
  label         = c("Baseline", "High\nTransmission", "Intervention"),
  color         = c("#2166ac", "#d6604d", "#4dac26")
)

p_r0 <- ggplot(r0_df, aes(x = R0, y = eq_prevalence)) +
  geom_line(linewidth = 1.2, color = "grey40") +
  geom_vline(xintercept = 1, linetype = "dashed", color = "grey60") +
  geom_point(data = sc_points, aes(color = color), size = 4, show.legend = FALSE) +
  geom_text( data = sc_points, aes(label = label, color = color),
             vjust = -0.8, size = 3.5, show.legend = FALSE) +
  scale_color_identity() +
  scale_x_continuous(breaks = seq(1.0, 3.0, by = 0.25)) +
  labs(
    title    = "SIS Model: Equilibrium Prevalence as a Function of R0",
    subtitle = "Deterministic analytical result; points show simulated scenarios",
    x        = "Basic Reproduction Number (R0)",
    y        = "Equilibrium Prevalence (%)",
    caption  = "Below R0=1: pathogen cannot sustain; above R0=1: endemic equilibrium"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title       = element_text(face = "bold"),
    panel.grid.minor = element_blank()
  )

p_r0
ggsave("output/figures/r0_sensitivity.png", plot = p_r0,
       width = 8, height = 5, dpi = 300)
```

---

*Simulation Complete*
